\chapter{Acknowledgements}

I am grateful to my promotors, Prof. Dr. Ir. Sofie Van Hoecke and Prof. Dr. Bram Vervisch, not just for promoting this master's dissertation, but also for sharing their expertise in their respective domains with me.
I am also grateful to Prof. Dr. Bram Vervisch and my tutor David Vander Mijnsbrugge for their regular feedback on my work and the new insights I have thanks to meetings with them.
I realize this was not an easy task for them, given the technical issues at the automotive supplier company's side and me not always being able to spend my time on this master's dissertation.
The delays due to these factors made my work progress at a slower pace than they desired.
It was motivating to see that they nevertheless supported me until the very end of my work.

This master's dissertation is not just the result of the research and engineering effort of my promotors, my tutor, and I.
We should not underestimate the impact of working full-time and studying full-time on my personal life.
I am therefore especially grateful to my parents and friends for putting up with the consequences of my decision to complete the advanced master's programme in general.
A warm apology and thank you in particular are in order for my girlfriend Anne, who has had to miss me due to my coursework for the advanced master's programme on far too many occasions.
I wish her all the best in completing her own master's programme in the coming years.

Dylan Meysmans

Machelen, 26 August 2022
